Trail Saving on Backtrack

### 2020-02-25 figures/scatter.tex
experiments

\documentclass{standalone}
\usepackage{tikz,pgfplotstable,pgfplots}
\pgfplotsset{compat=1.13}
\begin{document}
\begin{tikzpicture}
\begin{axis}[title=Trail Saving Running Time Per Instance,axis equal,ymin=0,xmin=0,
width=11cm,height=11cm,xlabel=original time (s),ylabel=trail saving time (s),axis x line=bottom,axis y line=left,xlabel near ticks,ylabel near ticks]
\addplot[only marks,mark size=1pt,mark=*,mark options={fill=black}]
	table[x=original,y=trail]
{test2.txt};
\addplot [red,samples at={0,1}] {x};
\draw [blue,dashed] (rel axis cs:0,0) -- (rel axis cs:1,1);
\end{axis}
\end{tikzpicture}
\end{document}

>>>>

\documentclass{standalone}
\usepackage{tikz,pgfplotstable,pgfplots}
\pgfplotsset{compat=1.13}
\begin{document}
\begin{tikzpicture}
\begin{axis}[title=Propagations Per Second in Cadical,axis equal,ymin=0,xmin=0,
width=11cm,height=11cm,xlabel=Baseline props/sec (in M),ylabel=Trail Saving props/sec (in M),axis x line=bottom,axis y line=left,xlabel near ticks,ylabel near ticks]
\addplot[only marks,mark size=1pt,mark=*,mark options={fill=black}]
	table[x=original,y=trail]
{scatter.txt};
\addplot [red,samples at={0,1}] {x};
\draw [blue,dashed] (rel axis cs:0,0) -- (rel axis cs:1,1);
\end{axis}
\end{tikzpicture}
\end{document}

### 2020-02-25 paper.tex
experiments

Chronological backtracking or partial backtracking instead allows a
reduction in the length of the backtrack by placing literals on the
trail out of decision level order. By reducing the length of the
backtrack the solver can keep more of its assignment trail
intact. This can save it from the work involved in reconstructing a
lot of its trail. Using \cbt is not a panacea however. Its application
must be limited for peak effectiveness. This indicates that it is
sometimes beneficial for the solver backtrack fully and redo its
trail, even if this takes more work. We will expand on why this might
be case below.

In this paper we present an alternative ``trail savings'' method where
by we save the backtracked part of the solver's trail and attempt to
use that information to make the solver's redescent more
efficient. Trail saving preserves the traditional invariants of the
SAT solver and its basic version is very simple to implement. It
allows the search to retain complete control over the order of
decisions, but helps make propagation faster. We develop some
enhancements to make the idea more effective, and demonstrate
experimentally that it performs as well as and sometimes better than
chronological backtracking. We also show that with the enhancements we
are able to improve the performance of state of the art solvers.

@@

A \sat solver's trail satisfies a number of conditions. However, in
this work we will need some additional flexibility in our definitions,
as we will be sometimes be working with trails that would never be
constructed by a \sat solver. Hence, we define a \emph{trail} to be a
sequence of literals each of which is either a \emph{decision} literal
or an \emph{implied} literal, and each of which has a
\emph{reason}. These two types of literals are distinguished by their
\emph{reasons}. Decision literals $d$ have a null reason,
$\reason{d} = \varnothing$. Implied literals $l$ have as a reason a
clause of the formula $\formula$, $\reason{l} =
C\in\formula$. (The clause $\reason{l}$ can be a learnt clause that has been
added to $\formula$).

@@

In CDCL solvers using standard conflict directed backtracking all
properties hold of the prefix of the solver's trail consisting of all
decisions levels but the deepest. The full trail might, however,
contain a conflict at its deepest level so is not necessarily conflict
free. The full trail might also not be propagation-complete, as unit
propagation at the deepest level is typically terminated early if a
conflict is found. It can further be noted that the first four
properties imply that if a clause $C$ is falsified at decision level
$k$, then $C$ must contain at least two literals at level $k$
(otherwise $C$ would have become unit at a prior level and then
satisfied by making its last unfalsified literal \true).

@@

\section{Experiments and Results}
For the version of cadical published in
\cite{DBLP:conf/sat/MohleB19}chronological backtracking gives solves 6
more problems and has a lower Par-2 score by a decent margin. When
adding trail saving and all enhancements to the default
non-chronological version of cadical, we end up solving 4 more problems
and the Par-2 score goes decreases by a margin not as good as that
produced by the non-chronological backtracking version. The trail saving
versions improve on the standard non-chronological backtracking but do
slightly worse than the chronological backtracking versions.

For the latest version of cadical that we downloaded as of January 1,
2020, chronological backtracking (set to default parameters)
interestingly makes the solver perform worse than standard
non-chronological backtracking does. This shows that not all solvers
will benefit from chronological backtracking; it depends on the solver's
mix of heuristics and may require fine-tuning the parameters associated
with chronological backtracking. However, after implementing trail
saving and enhancements on top of this version of cadical with
non-chronological backtracking, the solver performs roughly the same as
it does with standard non-chronological backtracking, and does not
suffer a slow down as chronological backtracking appears to.

For MapleSat, chronological backtracking makes a significant improvement
on MapleLCMDist both in terms of total number of problems solved and
Par-2 score. Similarly, trail saving implemented on top of MapleLCMDist
with non-chronological backtracking receives a similarly significant
improvement in terms of total instances solved and Par-2 score, and
actually outperforms chronological backtracking on both measures.

\begin{figure}
    \begin{tabular}{|l|l|l|l|l|l|}
      \hline
      & Total Solved & SAT & UNSAT & Par-2 (x 10\textasciicircum{}6 s) \\ \hline
      cadical                  & 532          & 314 &  218  & 3.220                             \\ \hline
      cadical\_chrono          & 525          & 308 &  217  & 3.269                             \\ \hline
      cadical\_trail           & 529          & 310 &  219  & 3.248                             \\ \hline
      cadical\_trail\_enhanced & 531          & 313 &  218  & 3.216                             \\ \hline
    \end{tabular}
    \caption{Table of results for cadical, version pulled from github as of January 1, 2020.}
\end{figure}

\begin{figure}
    \begin{tabular}{|l|l|l|l|l|l|}
      \hline
      & Total Solved & SAT & UNSAT & Par-2 (x 10\textasciicircum{}6 s) \\ \hline
      cadical                  & 492          & 289 &  203  & 3.5020                            \\ \hline
      cadical\_chrono          & 498          & 292 &  206  & 3.4400                            \\ \hline
      cadical\_trail           & 493          & 290 &  203  & 3.5228                            \\ \hline
      cadical\_trail\_enhanced & 496          & 292 &  204  & 3.4847                            \\ \hline
    \end{tabular}
    \caption{Table of results for cadical or "chrono", version published in \cite{DBLP:conf/sat/MohleB19}.}
\end{figure}

\begin{figure}
    \begin{tabular}{|l|l|l|l|l|l|}
      \hline
      & Total Solved & SAT & UNSAT & Par-2 (x 10\textasciicircum{}6 s) \\ \hline
      maple           & 458          & 259 &  199  & 3.797                             \\ \hline
      maple\_chrono   & 470          & 271 &  199  & 3.690                             \\ \hline
      maple\_trail    & 472          & 271 &  201  & 3.694                             \\ \hline
      maple\_enhanced & 473          & 277 &  195  & 3.667                             \\ \hline
    \end{tabular}
    \caption{Table of results for MapleLCMDist.}
\end{figure}
\clearpage

>>>>

Chronological backtracking or partial backtracking instead allows a
reduction in the length of the backtrack by placing literals on the
trail out of decision level order. By reducing the length of the
backtrack the solver can keep more of its assignment trail
intact. This can save it from the work involved in reconstructing a
lot of its trail. Using \cbt is not a panacea however. Its application
must be limited for peak effectiveness. This indicates that it is
sometimes beneficial for the solver to backtrack fully and redo its
trail, even if this takes more work. We will expand on why this might
be the case below.

In this paper we present an alternative ``trail savings'' method whereby
we save the backtracked part of the solver's trail and attempt to
use that information to make the solver's redescent more
efficient. Trail saving preserves the traditional invariants of the
SAT solver and its basic version is very simple to implement. It
allows the search to retain complete control over the order of
decisions, but helps make propagation faster. We develop some
enhancements to make the idea more effective, and demonstrate
experimentally that it performs as well as and sometimes better than
chronological backtracking. We also show that with the enhancements we
are able to improve the performance of state-of-the-art solvers.

@@

A \sat solver's trail satisfies a number of conditions. However, in
this work we will need some additional flexibility in our definitions,
as we will sometimes be working with trails that would never be
constructed by a \sat solver. Hence, we define a \emph{trail} to be a
sequence of literals each of which is either a \emph{decision} literal
or an \emph{implied} literal, and each of which has a
\emph{reason}. These two types of literals are distinguished by their
\emph{reasons}. Decision literals $d$ have a null reason,
$\reason{d} = \varnothing$. Implied literals $l$ have as a reason a
clause of the formula $\formula$, $\reason{l} =
C\in\formula$. (The clause $\reason{l}$ can be a learnt clause that has been
added to $\formula$).

@@

In CDCL solvers using standard conflict directed backtracking all
properties hold of the prefix of the solver's trail consisting of all
decisions levels but the deepest. The full trail might, however,
contain a conflict at its deepest level so is not necessarily conflict-
free. The full trail might also not be propagation-complete, as unit
propagation at the deepest level is typically terminated early if a
conflict is found. It can further be noted that the first four
properties imply that if a clause $C$ is falsified at decision level
$k$, then $C$ must contain at least two literals at level $k$
(otherwise $C$ would have become unit at a prior level and then
satisfied by making its last unfalsified literal \true).

@@

\section{Experiments and Results}
We implemented our techniques in two different SAT solvers, MapleSAT and Cadical,
both of which have finished at or near the top of SAT competitions for the past several
years \cite{Heule2019ProceedingsOS}. We then ran each solver on the 800 benchmark instances used
in the main tracks of the 2018 and 2019 SAT competitions. The
experiments were executed on a cluster of 2.7 GHz Intel cores with 5000 seconds CPU time and 7 GB memory limits for each instance. We chose not to output or verify the proofs generated by any of the solvers. The Par-2 scores obtained and total instances solved by each solver are reported in Figures 3, 4, and 5.\newline

In Figure 3, we use the newest version of cadical (downloaded as of January 1, 2020) as the baseline solver. In our table, cadical refers to the baseline solver with non-chronological backtracking, cadical-chrono refers to the solver with \cbt enabled (using default settings), cadical-trail is the baseline with plain trail saving added (as described in Figure~\ref{fig:trailUse}), cadical-trail-append is the baseline with trail saving plus the first enhancement of saving over multiple backtracks, cadical-trail-append-lookahead adds the enhancement of lookahead for conflicts by 2 decision levels, and lastly cadical-trail-append-lookahead-reason also adds the final enhancement to cease trail saving once a reason of size greater than 10 literals is reached. For more details on the enhancments, please see section 4.2.

Interestingly, \cbt makes the solver perform worse than the baseline. However, after implementing trail saving and all enhancements, the solver outperforms the baseline by a decent margin. The key enhancement for this solver seems to be the last one where we stop using the saved trail once we detect a reason of "poor" quality. We tried both clause size and lbd to act as a measure of clause quality, and both yielded a positive gain, with clause size being slightly more effective. \newline

In Figure 4, we use the version of cadical published in
\cite{DBLP:conf/sat/MohleB19} as a baseline, since we know it is a version of cadical that \cbt works particularly well on. When
adding trail saving and all enhancements to the baseline, it solves the same number of instances as
\cbt. Both \cbt and trail saving boost this solver's performance significantly. MapleLCMDist is another solver we know \cbt works well on, and in Figure 5 we use it as the baseline solver. \cbt and trail saving both make a significant improvement on this solver, but trail saving performs slightly better than \cbt in terms of total instances solved and Par-2 score.\newline

In Figure 6 we plot the propagations per second for each instance with and without trail saving. With trail saving about 1.87 million literals per second are propagated on average, and without trail saving about 1.80 million literals per second are propagated on average. The graph further illustrates that the majority of the time trail saving will propagate literals faster than a solver without trail saving, as we argued earlier in the paper.

\begin{figure}
\centering
    \begin{tabular}{|l|l|l|l|l|l|}
      \hline
      & Total & SAT & UNSAT & Avg. Par-2 \\ \hline
      cadical                  & 532          & 314 &  218  & 4025                             \\ \hline
      cadical-chrono          & 525          & 308 &  217  & 4086                             \\ \hline
      cadical-trail           & 529          & 310 &  \textbf{219}  & 4060                   \\ \hline
      cadical-trail-append & 530 & 313 & 217 & 3930 \\ \hline
      cadical-trail-append-lookahead & 531          & 313 &  218  & 4020                      \\ \hline
      cadical-trail-append-lookahead-reason & \textbf{538} & \textbf{319} & \textbf{219} & \textbf{3854} \\ \hline
    \end{tabular}
    \caption{Table of results for cadical, version pulled from github as of January 1, 2020.}
\end{figure}

\begin{figure}
\centering
    \begin{tabular}{|l|l|l|l|l|l|}
      \hline
      & Total & SAT & UNSAT & Avg. Par-2 \\ \hline
      cadical                  & 492          & 289 &  203  & 4378                            \\ \hline
      cadical-chrono   & \textbf{498}  & \textbf{292} &  \textbf{206}  & \textbf{4300}      \\ \hline
      cadical-trail           & 493          & 290 &  203  & 4403                            \\ \hline
      cadical-trail-append-lookahead & 496  & \textbf{292} &  204  & 4356                    \\ \hline
      cadical-trail-append-lookahead-reason & \textbf{498} & \textbf{292} & \textbf{206} & 4351 \\ \hline
    \end{tabular}
    \caption{Table of results for cadical or "chrono", version published in \cite{DBLP:conf/sat/MohleB19}.}
\end{figure}

\begin{figure}
\centering
    \begin{tabular}{|l|l|l|l|l|l|}
      \hline
      & Total & SAT & UNSAT & Avg. Par-2 \\ \hline
      maple           & 458          & 259 &  199  & 4746                             \\ \hline
      maple-chrono   & 470          & 271 &  199  & 4613                             \\ \hline
      maple-trail    & 472          & 271 &  \textbf{201}  & 4618                \\ \hline
      maple-trail-append-lookahead & \textbf{473} & \textbf{277} &  195  & \textbf{4584} \\ \hline
    \end{tabular}
    \caption{Table of results for MapleLCMDist.}
\end{figure}

\begin{figure} \centering \includegraphics[scale=0.7]{figures/scatter.pdf}\caption{Comparing propagations per second with and without trail saving in Cadical.}\end{figure}
